% Compile: pdflatex dynamics_models.tex  (self-contained; no external .bib needed)
\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{bm}
\usepackage[hidelinks]{hyperref}

\newcommand{\q}{\bm{q}}
\newcommand{\qd}{\dot{\bm{q}}}
\newcommand{\qdd}{\ddot{\bm{q}}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\M}{\bm{M}}
\newcommand{\Lo}{\bm{L}}

\title{Physics-Structured Dynamics Models for Model-Based RL\\
\large Lagrangian formulation, network parameterization, and the controller}
\author{lagrangian-mbrl-franka}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
We learn the forward dynamics of the 7-DoF Franka Emika Panda,
$(\q,\qd,\bm{\tau})\mapsto\qdd$, to use as the predictive model inside a
model-based reinforcement-learning (MBRL) loop. This note states the
Euler--Lagrange equations the dynamics obey, shows \emph{exactly where the
neural network enters} a Deep Lagrangian Network (DeLaN), lists the architectures
used for the structured model and the unstructured baselines, and specifies the
controller that plans through the learned model.
\end{abstract}

\section{Problem}
A rigid-body manipulator with generalized coordinates $\q\in\R^{n}$
($n=7$ for the Franka) and applied joint torques $\bm{\tau}\in\R^{n}$ evolves
according to the manipulator equation
\begin{equation}
\M(\q)\,\qdd + \bm{c}(\q,\qd) + \bm{g}(\q) = \bm{\tau},
\label{eq:manip}
\end{equation}
where $\M(\q)\succ0$ is the (symmetric positive-definite) mass/inertia matrix,
$\bm{c}$ collects Coriolis/centrifugal terms, and $\bm{g}$ is the gravitational
force. The learning target is the \emph{forward dynamics}
\begin{equation}
\qdd = \M(\q)^{-1}\!\left(\bm{\tau}-\bm{c}(\q,\qd)-\bm{g}(\q)\right).
\label{eq:fwd}
\end{equation}

\section{Lagrangian mechanics}
The dynamics \eqref{eq:manip} are the Euler--Lagrange equations of the Lagrangian
$\mathcal{L}=T-V$, with kinetic energy $T=\tfrac12\qd^{\top}\M(\q)\qd$ and
potential $V(\q)$:
\begin{equation}
\frac{d}{dt}\!\left(\frac{\partial \mathcal{L}}{\partial \qd}\right)
-\frac{\partial \mathcal{L}}{\partial \q} \;=\; \bm{\tau}.
\label{eq:el}
\end{equation}
Carrying out the differentiation recovers \eqref{eq:manip} with
$\bm{g}(\q)=\partial V/\partial\q$ and the Coriolis term expressed through
derivatives of $\M$. Encoding this structure is the whole point: the true
dynamics \emph{lie in} this class, so a model constrained to it has a smaller
effective hypothesis space and needs fewer samples to reach a given accuracy
(the sample-complexity argument of \texttt{PROJECT\_PLAN.md}~\S2).

\section{Where the network goes: Deep Lagrangian Networks (DeLaN)}
\label{sec:delan}
DeLaN \cite{lutter2019delan} does \emph{not} regress $\qdd$ directly. Instead,
two neural networks parameterize the \emph{energy}, and the dynamics follow by
automatic differentiation of \eqref{eq:el}. Concretely, with parameters
$\bm{\theta}$:
\begin{itemize}
  \item a network outputs the entries of a lower-triangular
  $\Lo(\q;\bm{\theta})$, giving a guaranteed-SPD mass matrix
  \begin{equation}
  \M(\q;\bm{\theta}) = \Lo(\q;\bm{\theta})\,\Lo(\q;\bm{\theta})^{\top}
  + \varepsilon\bm{I}, \qquad \varepsilon>0,
  \label{eq:cholesky}
  \end{equation}
  (the diagonal of $\Lo$ is passed through a shifted softplus so it stays
  strictly positive, which conditions the solves);
  \item a scalar network outputs the potential $V(\q;\bm{\theta})$.
\end{itemize}
The kinetic-plus-potential Lagrangian
$\mathcal{L}=\tfrac12\qd^{\top}\M(\q;\bm\theta)\qd-V(\q;\bm\theta)$ is then
differentiated through \eqref{eq:el}. Using the identity implemented in the code,
the combined Coriolis-plus-gravity force is
\begin{equation}
\bm{c}(\q,\qd)+\bm{g}(\q) \;=\;
\left(\frac{\partial^{2}\mathcal{L}}{\partial\qd\,\partial\q}\right)\!\qd
\;-\;\frac{\partial \mathcal{L}}{\partial \q},
\label{eq:cg}
\end{equation}
and the predicted acceleration follows from \eqref{eq:fwd} by solving the SPD
system $\M\,\qdd = \bm\tau-\bm c-\bm g$ (a Cholesky solve). Because
\eqref{eq:cg} takes \emph{second} derivatives of $\mathcal{L}$, the network
activations must be twice continuously differentiable ($C^{2}$): we use
\textbf{softplus} (or \textbf{tanh}), never ReLU.

\paragraph{Training.}
We minimize a one-step loss on $(\q,\qd,\bm\tau,\qdd)$ transitions combining the
forward-dynamics residual (predicted vs.\ measured $\qdd$) and the
inverse-dynamics residual (predicted vs.\ measured $\bm\tau$), with input
normalization fit on the training set.

\section{Model architectures (what we use in each case)}
\label{sec:arch}
\begin{table}[h]
\centering
\begin{tabular}{@{}lll@{}}
\toprule
Role & Model & Architecture \\
\midrule
Structured (\emph{proposed}) & DeLaN \cite{lutter2019delan} &
  energy heads $\Lo(\q),V(\q)$, MLP $[128,128]$, \textbf{softplus} ($C^2$),
  SPD via \eqref{eq:cholesky} \\
Unstructured baseline & MLP ensemble (PETS) \cite{chua2018pets} &
  $K{=}5$ probabilistic MLPs $[256,256,256]$, SiLU, Gaussian $\qdd$ head \\
Minimal reference & MLP & single deterministic MLP $[256,256,256]$, SiLU \\
\bottomrule
\end{tabular}
\caption{Models compared. The structured model constrains the hypothesis class
to physically consistent dynamics; the ensemble is the strong unstructured
baseline and supplies epistemic uncertainty for planning. Widths/activations are
the defaults; the benchmark tunes them per model.}
\end{table}

\paragraph{Related structured models (future ablations).}
Lagrangian Neural Networks \cite{cranmer2020lnn} learn an \emph{arbitrary}
$\mathcal{L}$ (dropping the $T=\tfrac12\qd^\top\M\qd$ assumption); Hamiltonian
Neural Networks \cite{greydanus2019hnn} parameterize the Hamiltonian instead;
and Lagrangian Graph Neural Networks \cite{bhattoo2022lgnn} exploit the
manipulator's kinematic topology and scale to many links. They slot into the
same model registry as drop-in alternatives to DeLaN.

\section{The controller: planning through the learned model}
\label{sec:control}
Given the learned forward model $\hat f_{\bm\theta}:(\q,\qd,\bm\tau)\mapsto\qdd$
and a one-step integrator $\hat F$ producing $(\q_{t+1},\qd_{t+1})$, control is
\emph{model-predictive}. At each state $\bm{x}_t=(\q_t,\qd_t)$ we optimize a
torque sequence over a horizon $H$,
\begin{equation}
\min_{\bm\tau_{t:t+H-1}} \;
\sum_{k=0}^{H-1}\ell(\bm{x}_{t+k},\bm\tau_{t+k}) + \ell_f(\bm{x}_{t+H}),
\qquad
\bm{x}_{t+k+1}=\hat F(\bm{x}_{t+k},\bm\tau_{t+k}),
\label{eq:mpc}
\end{equation}
apply the first torque, and re-plan (receding horizon). The default solver is
\textbf{MPPI} \cite{williams2017mppi}: sample $N$ noisy control sequences, roll
each through $\hat F$, and update the control by the cost-weighted average
\begin{equation}
\bm\tau \;\leftarrow\; \sum_{i=1}^{N} w_i\,\bm\tau^{(i)},
\qquad
w_i \propto \exp\!\Big(-\tfrac1\lambda\,\big(J^{(i)}-\min_j J^{(j)}\big)\Big),
\label{eq:mppi}
\end{equation}
with $J^{(i)}$ the rolled-out cost \eqref{eq:mpc} and temperature $\lambda$.
MPPI needs \emph{no policy network}, so its performance is governed almost
entirely by the accuracy of $\hat f_{\bm\theta}$---this is the cleanest probe of
whether the Lagrangian prior helps. The reach cost is the end-effector position
(and orientation) error to the commanded pose plus a torque-effort penalty.

\paragraph{Alternatives.}
A Dyna/MBPO variant \cite{janner2019mbpo} would instead generate short model
rollouts and run an off-policy update (e.g.\ SAC) on mixed real+model data;
a model-free PPO policy serves as the lower-bound baseline. These are deferred
until MPPI results justify a learned policy.

\section{Evaluation protocol}
Models are compared on held-out one-step acceleration MSE (and, for the
structured model, energy drift and $\lambda_{\min}(\M)$), under matched data:
the same seeds and the same per-seed transitions feed every model. We tune
hyper-parameters per model, retrain the winner over $\ge 5$ seeds, and report the
mean with a percentile-bootstrap 95\% confidence interval. The data-efficiency
story is the curve of model error (and, later, task return) versus the number of
real transitions $N$.

\begin{thebibliography}{9}
\bibitem{lutter2019delan} M.~Lutter, C.~Ritter, J.~Peters.
  \emph{Deep Lagrangian Networks: Using Physics as Model Prior for Deep
  Learning.} ICLR, 2019.
\bibitem{cranmer2020lnn} M.~Cranmer et al.
  \emph{Lagrangian Neural Networks.} ICLR Deep Differential Equations
  Workshop, 2020.
\bibitem{greydanus2019hnn} S.~Greydanus, M.~Dzamba, J.~Yosinski.
  \emph{Hamiltonian Neural Networks.} NeurIPS, 2019.
\bibitem{bhattoo2022lgnn} R.~Bhattoo, S.~Ranu, N.~M.~A.~Krishnan.
  \emph{Learning Articulated Rigid Body Dynamics with Lagrangian Graph Neural
  Network.} NeurIPS, 2022.
\bibitem{chua2018pets} K.~Chua, R.~Calandra, R.~McAllister, S.~Levine.
  \emph{Deep Reinforcement Learning in a Handful of Trials using Probabilistic
  Dynamics Models (PETS).} NeurIPS, 2018.
\bibitem{janner2019mbpo} M.~Janner, J.~Fu, M.~Zhang, S.~Levine.
  \emph{When to Trust Your Model: Model-Based Policy Optimization (MBPO).}
  NeurIPS, 2019.
\bibitem{williams2017mppi} G.~Williams, A.~Aldrich, E.~A.~Theodorou.
  \emph{Model Predictive Path Integral Control.} Journal of Guidance, Control,
  and Dynamics, 2017.
\end{thebibliography}

\end{document}
